Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-11-08 analysis/paper/Letter_Response_to_Reviewers.qmd

::: author-response
**\[author response\]** We address this in the manuscript by introducing the importance of the statutory requirements of the MMPA and ESA early in the second paragraph of the introduction. We hope this frames the important link in the mind of the reader between the work described here and the need for improved abundance estimates.

Regarding the desire for specific examples of how the modeled availability would have impacted population abundance estimates from aerial survey data, we don't have the ability to provide such an example because the aerial survey data are not yet published and available. Such an example and discussion might be better suited for subsequent papers that focus on presentation of the aerial survey data and the abundance estimation process.
:::

> *The findings are appear robust and intuitive. The authors need to improve the presentation of the findings to improve clarity and help the reader understand how they addressed the knowledge gap.*

::: author-response
**\[author response\]** The reviewer provided numerous comments and suggestions annotated throughout the PDF that address the concern of presentation and clarity. In the next section we provide responses to those annotations
:::

@@

> ***Title:** The reviewer suggests removing and aerial survey detectability from the title of the manuscript and suggests adding the seasonal time period our study focused on.*

::: author-response
**\[author response\]** We have changed the title to: Spring haul-out behavior of seals in the Bering and Chukchi seas
:::

> ***Abstract:** Can you make a statement in your abstract that use of modeled haulout behavior using covariates available from regional reananalysis data improved the confidence intervals on availability estimates and that this would improve aerial survey abundance estimation efforts?*

::: author-response
**\[author response\]** We don't feel comfortable making a precision comparison with previous abundance estimates, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters). As such, it will be difficult to ascribe a specific reason to precision increases or decreases.
:::

@@

> *Please use direct language. You are not making findings on how seals rely on sea ice, rather you are improving methods to estimate their availability for improved aerial survey population estimation efforts. Here you are focused on modeling ice sea availability to aerial survey so that aerial survey counts may be appropriately interpreted. Please be clear that you are considering ice seal behaviors that may be telemtered as "dry out of water behaviors" that may be interpreted as behaviors that place the seal on top of the ice and available for observation by an aerial survey.*

::: author-response
**\[author response\]** The opening sentence of the abstract is a statement regarding the reliance of ice-associated seals on sea-ice for a range of important life history behaviors. We think this statement is an important introduction for readers who are not familiar with ice-associated seals. We do agree with the reviewer's suggestion for more direct language and have made edits to the beginning of the abstract with this in mind.
:::

@@

::: author-response
**\[author response\]** This sentence now reads:

"Ship-based capture events occurred during spring near the southern ice edge in the Bering Sea between 2005 and 2018."
:::

> ***Line 124:** Please provide the geographic context of your tag deployments and data collection using a plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 2). Please consider using a tabulation that enables readers to understand the geographic context of the deployments similar to that used by Jay et al. 2012, table 1.*

::: author-response
**\[author response\]** We feel the spatial distribution of the actual transmitted data -- compared to the deployment location -- is most informative and relevant for the audience's understanding of how our data are distributed spatially and how well data represents the range of each species. We provide this visually in Figure 3 and adding an additional layer to that figure to show deployments would be an additional complication. To ensure the audience has a clear understanding of our study area, we have added an additional figure (now, Figure 1) with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated. In response to other reviewer comments this figure also indicates the spatial range of aerial survey effort to date.
:::

@@

> ***Line 224:** Using the term "major focus" implies that there may be numerous minor foci. Can you not state that this is the goal of the study? Ancillary natural history insights can be discussed incidental to this, but are not driving the research. The statutory requirements of the MMPA and ESA are driving your study. These Federal laws require robust methods to monitor population status. Your study seeks to build this method. Discussion of natural history belong in your Discussion section and not in the framing of your study. Please help the reader focus on why the Federal government is doing this study.*

> *Understanding pinniped haulout behavior is essential to designing and interpreting population estimation aerial surveys. Understanding how they use sea ice during a time when sea ice availability is rapidly changing is also important, but is ancillary to the paper's focus.*

::: author-response
**\[author response\]** We have removed the term 'major focus' and, as the reviewer suggests, simply state that this is the goal of the study. We have also added mention of the statutory requirements to paragraph 2 of the Introduction section and expanded the text to acknowledge the importance of understanding how seals use sea ice during this period of change but, ultimately, it's ancillary to the focus of this paper.

As for providing the reader focus on why the Federal government is doing this study: this is a collaborative study across multiple organizations and the analysis/purpose is not limited to the Federal government. We've incorporated the reviewers suggestions to expand discussion of the statutory requirements of the MMPA and ESA but don't feel we need to single out the motivations for the Federal government.

We have also added additional general representation of recent aerial survey effort to the new Figure 1. This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
:::

@@

> ***Line 227:** Seals can just as happily be on foraging trips within the pack and marginal sea ice as in the open water, depending on the species. Suggesting that foraging trips are associated with ice-free conditions is misleading.*

::: author-response
**\[author response\]** We have removed 'e.g., foraging trips' from this sentence so as not to mislead the reader.
:::

> ***Line 229-230:** This is your primary reason for not including sea ice geography in your analysis: you simply lacked geolocation of much of your haulout behavior data.*
>
> *Please help the reader understand how coarse your tracking data was for the flipper tagged animals and what proportion of each species was tracked with these flipper tags. Please also help the reader understand how the tracking data relates geographically to regions that may be subject to aerial survey efforts (i.e. are your tracking data collected from across the regions that may be surveyed during range-wide spring aerial surveys?).*

::: author-response
**\[author response\]** We have re-organized this paragraph to, hopefully, better communicate our choice not to explore sea-ice related covariates. As the reviewer suggests, we have limited geo-location data for at-sea periods of our haul-out behavior data -- we know the seal was in the water but don't have any estimates of location when only a flipper tag is attached. By leading with this statement, we hope that communicates our primary rationale. But, the other rationales are also important and we still include them. We have also moved this section to the 'Discussion' in response to other reviewer comments.

@@

We have provided some statistics regarding the percentage of the percent-dry timeline data that is not associated with a corresponding location estimate because the seal was in the water and no estimates were available from flipper mounted SPOT tags.

We have also added a figure 1 depicting a spatial representation of recent aerial survey effort (which reflects estimated species range for bearded, ribbon, and spotted seals in US waters). This should better communicate to the reader how the spatial distribution of our haul-out behavior overlaps with the species range and survey effort.
:::

@@

> ***Line 265-272:** Please seek input from a competent biometrician. I am not able to offer any guidance on this other than to muse why you are pushing to estimate six parameters for a time of day covariate when your seals are not known to keep track of time.*
>
> *It would be reasonable to assume that seals respond to insolation or the angular height of the sun in its daily arc.  Please help your reader understand why you have sought to model a complex representation of hourly time when you could have pulled insolation from NARR or calculated sun angular altitude (perhaps as a linear proportion of the day's maximum sun angular altitude) from standard astronomical functions.*
>
> *Perhaps your motivation is to provide plots that appear intuitive with hours on the y-axis, day of year on the x-axis and haulout proportion as a "heat map".  However beautiful these plot are, I wonder if your goal for improving precision and reducing bias when estimating availability during an aerial survey would not be better served by  the alternative and simpler variables.*

::: author-response
**\[author response\]** We appreciate the suggestion and find the possibility of using insolation values from reanalysis products as a covariate intriguing. As the reviewer suggests, such an approach could provide a less complicated and more parsimonious model than the approach we have taken with the Fourier series as a cyclic representation of hour-of day. We have taken the time to do some additional analysis and methodological exploration and provide additional details in Supplement 2. Note, our initial exploration of this approach resulted in lower AIC values for the Fourier series approach presented in the analysis. While promising and worth even further exploration, we feel the approach taken in the analysis is the most appropriate because it distinguishes daily bi-modal (e.g. crepuscular) patterns that might be important but would not be detected by insolation alone (i.e. insolation at dawn or dusk would be similar values but occur at different times of day that might be ecologically important). Additionally, as the reviewer points out, results from the analysis and haul-out predictions are more intuitive and human interpretable when using hour-of-day.
:::

@@

> ***Line 278:** Please help your reader understand what is meant by your term "horizontal shift".  It is clear that you have a graphical representation in your mind that has not been shared with your reader. Please rewrite to avoid "jargon".*

> ***Line 285:** Please help your reader understand what could be "important" about these terms. Please write for clarity, avoiding jargon.*

> ***Figure 4:** Please use standard conventions for plot titles.  Your title deserves a statement in your Results or Discussion section.  It does not follow the expectations for a publication figure title.*

::: author-response
**\[author response\]** The plot titles were developed with strong consideration for improving scientific communication over approaches that are more typically used in publications. They were developed with guidance from modern approaches to data visualization that recognize the plot title is often the very first thing the reader sees and that data visualizations are often viewed and interpreted as standalone figures. Therefore it can serve as an effective communication tool to help the reader interpret the results.

That said, we recognize that the initial approach could still be improved and have made adjustments to each of the titles with the feedback in mind. We hope the new titles provide a balanced approach.
:::

@@

> ***Figure 4:** Please scale your color ramp to enable better visualization of the subject: limit the color ramp to the modeled values (which appear to be from zero to \~0.5).*

::: author-response
**\[author response\]** The scale used is consistent across all three species to maintain comparability. If we were to adjust the scale as suggested it would make comparisons across species difficult
:::

> ***Figure 4:** Please switch to a color ramp that is suitable for readers with color perception disabilities such as the viridis color ramps.*

::: author-response
**\[author response\]** The color ramp used has been updated to use palettes from the R package 'scico' have been tested and confirmed to be colorblind friendly and appropriately balanced for scientific interpretation. We initially used the viridis color palette for these plots but the 'scico' palette colors provided additional visual detail we feel is important.
:::

### 2023-11-08 analysis/paper/London_HauloutBeahvior.Rmd
major update to qmd and figures

  between 2005 and 2021 to quantify the proportion of seals hauled out on ice,
  which is needed to accurately estimate abundance from aerial survey counts of

@@

  strong diurnal and within-season patterns in haul-out behavior, as well as

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005a], [@conn2014a], and [@verhoef2014a]) and more recent efforts
have significantly expanded on previous survey effort. Such abundance studies
are conducted over very large areas and estimation of absolute abundance
requires making inference about numerous issues affecting the observation of
seals on ice. These include availability (only seals on ice are available to
be counted), detection probability (observers or automated detection systems may
miss some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2014a; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

@@

The remoteness of the Bering and Chukchi seas means direct scientific
observation of seal behavior is impractical. Thus, bio-logging devices are
especially useful tools for collecting key information on movement and haul-out
behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in
haul-out behavior. For instance,Von Duyke et al. [-@vonduyke2020] used
satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
nocturnal haul-out behavior of ringed seals previously described by Kelly and
Quakenbush [-@kelly1990c] using VHF radio tags and direct observation. Bengtson
et al. [-@bengtson2005a] documented a higher propensity for ringed seal basking
near solar noon, as did Ver Hoef et al. [-@verhoef2014a] in an analysis of
bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
[-@olnes2020] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer, and Ver Hoef et al.
[-@verhoef2014a] found haul-out probabilities increased gradually starting in
March and peaked in May and June for bearded, ribbon, and spotted seals. Such
analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
Cameron [-@bengtson2004] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (_Lobodon carcinophaga_) than
adults, with highest probabilities in February and at times close to solar noon.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a])
accounted for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
variability in haul-out timing and possible linkage to changes in the extent of
seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability, and
investigates the potential impact of changing sea-ice extent on the behavior of
these species.

@@

Haul-out behavior data were recorded in a manner standard across Wildlife Computers
bio-loggers and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
inclusive of 0% and 100% along with additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Memory capacity allowed caching of percent-dry
records for several weeks or months and each message was transmitted several
times to ensure reception at the satellite. Bio-loggers were, typically,
deployed and programmed in a manner to maximize data transmission during the
spring pupping and molting period, though hourly percent-dry data were not
always successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
atmospheric interference. Fortunately, missing records do not substantially bias
inference about haul-out probabilities [@conn2012b].

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering and Chukchi seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

For all seal species, we considered the following variables when modeling the
hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
barometric pressure, precipitation, and wind chill (represented by a
*wind:temperature* interaction [@udevitz2009]). Ribbon and spotted seal models
included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of
age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
latitudinal effect (and an interaction with day-of-year), since bearded seals
occupy a substantial range during the spring and we were interested in possible
differences in the timing of haul-out behavior along a latitudinal gradient. We
omitted the latitudinal effect from ribbon and spotted seal models because,
during the spring, these species are most prevalent near the southern ice edge
in the Bering Sea [@conn2014a].

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal); see also 
@\ref(fig:beardedPredSE)). Additionally, after
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data (100% dry for several weeks but
indicative of movement and increasing transmission rates (see @boveng2013)). All
data after the first instance of unrealistic values were censored from this
analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
distribution of haul-out probability across the day. In addition to a peak
around local solar noon, the bearded seal model predicted additional haul-out
activity around local midnight. In concert with the lower magnitude of haul-out
probability, bearded seal haul-out behavior was also more protracted throughout
the spring season compared to ribbon and spotted seals.

@@

When exploring the influence of weather, bearded seal haul-out probability
appeared most influenced by wind
($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
and temperature
($F_{`r paste(bearded_type3$temp2$Num.df, bearded_type3$temp2$Den.df, sep=",")`}$
= `r bearded_type3$temp2$F.value`; $p$ =
`r bearded_type3$temp2$Prob.F`) (Figure \@ref(fig:beardedHOwx)). 
Additionally, wind chill was also a significant influence
($F_{`r paste(bearded_type3$temp2_wind$Num.df, bearded_type3$temp2_wind$Den.df, sep=",")`}$
= `r bearded_type3$temp2_wind$F.value`; $p$ =
`r bearded_type3$temp2_wind$Prob.F`). Any influence of barometric
pressure or precipitation was less apparent.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal);
see also @\ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
in April, May and June hauled out. In contrast to ribbon seals, adult spotted
seal males had a more protracted haul-out season compared to females, and more
time out of the water in June (Figures \@ref(fig:ribbonHOCal) and
\@ref(fig:spottedHOCal)). As with ribbon seals, the young-of-the-year records
began after weaning and the model predictions demonstrated the development of
in-water activities (e.g. swimming, foraging) in May.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2014a]), we also investigated the influence of
sex-age class on haul-out probabilities for both species (not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

@@

m2_ssrd_bearded <- mgcv::bam(
  dry ~ s(speno, bs = "re") + 
    era5_ssrd +
    poly(day, 3, raw=TRUE) + 
    era5_ssrd:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip,
  data = bearded_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_bearded,
  discrete = TRUE)
```

>>>>

  between 2005 and 2021 to quantify the proportion of seals hauled out on ice.
  This information is needed to accurately estimate abundance from aerial survey counts of

@@

  strong diel and within-season patterns in haul-out behavior, as well as

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2018]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting
[@boveng2009; @boveng2013a; @cameron2010; @kelly2010]. Limited data and large
knowledge gaps complicate predictions about the ultimate effects of changes in
sea ice on the behavior, health, abundance, and distribution of these seals. To
date, indices of seal health sampled during periods of declining sea ice differ
regionally [@crawford2015; @harwood2020]. Knowledge about evolutionary
constraints on the timing of reproductive and molting behavior is generally
lacking, so it is difficult to predict how or if ice-associated seal species
adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). This is further complicated by the
spatio-temporal variation in the phenology of these life history events within
regions and throughout their full ranges. Additionally, trends in abundance of
these species are unknown, so it is difficult to assess the effect, if any,
declines in sea ice habitat have had, or will have, on seal demography.

Statutory requirements (e.g., United States Marine Mammal Protection Act (MMPA),
United States Endangered Species Act (ESA)) for timely estimates of population
abundance and trend of these species mean improved aerial survey effort is
needed -- and, must be paired with improved knowledge of haul-out behavior to
ensure appropriate survey design, robust methods, and accurate estimates.
Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005a], [@conn2013b], and [@verhoef2014a]) and more recent efforts
have significantly expanded on previous survey effort. Such abundance studies
are conducted over very large areas and estimation of absolute abundance
requires making inference about numerous issues affecting the observation of
seals on ice. These include availability (only seals on ice are available to
be counted), detection probability (observers or automated detection systems may
miss some seals on ice), species misclassification, and possible
disturbance of seals by aircraft [@conn2013b; @verhoef2014a]. Refining these
inferences will improve the accuracy of abundance estimates and, hopefully,
allow credible predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations.

@@

The remoteness of the Bering and Chukchi seas means direct scientific
observation of seal behavior is impractical. Thus, bio-logging devices are
especially useful tools for collecting key information on movement and haul-out
behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in
haul-out behavior. For instance,Von Duyke et al. [-@vonduyke2020] used
satellite-linked bio-loggers to corroborate seasonal changes between diurnal and
nocturnal haul-out behavior of ringed seals previously described by Kelly and
Quakenbush [-@kelly1990c] using VHF radio tags and direct observation. Bengtson
et al. [-@bengtson2005a] documented a higher propensity for ringed seal basking
near solar noon, as did Ver Hoef et al. [-@verhoef2014a] in an analysis of
bearded, ribbon, and spotted seals using much larger sample sizes. Olnes et al.
[-@olnes2020] showed that the proportion of time bearded seals spent hauled out
progressively increased through spring and summer, and Ver Hoef et al.
[-@verhoef2014a] found haul-out probabilities increased gradually starting in
March and peaked in May and June for bearded, ribbon, and spotted seals. Recent
Such
analyses have not been limited to the Arctic. In the Antarctic, Bengtson and
Cameron [-@bengtson2004] relied on bio-logging data to demonstrate greater
haul-out propensity in juvenile crabeater seals (_Lobodon carcinophaga_) than
adults, with highest probabilities in February and at times close to solar noon.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], [@southwell2008], and [@niemi2023]). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we sought to refine
estimates of haul-out availability corrections for aerial surveys in order to
improve estimates of seal abundance. Previously estimated availability
correction factors (e.g., [@bengtson2005a], [@conn2013b], and [@verhoef2014a])
accounted for variables such as the time-of-day and day-of-year, but did not
investigate meteorological variables that have been shown to influence haul-out
behavior of walruses [@udevitz2009]. Third, we aimed to assess the annual
variability in haul-out timing and possible linkage to changes in the extent of
seasonal sea ice between 2005 and 2020. Our work extends the scope of previous
haul-out analyses, includes the influence of meteorological variability, and
investigates the potential impact of changing sea-ice extent on the behavior of
these species.

@@

Haul-out behavior data were recorded in a manner standard across Wildlife Computers
bio-loggers and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and the percent of the
hour in the dry state was calculated (Figure \@ref(fig:examplePlot)). On board
the bio-logger, hourly percent-dry calculations were rounded to the nearest 10%
inclusive of 0% and 100% along with additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Memory capacity allowed caching of percent-dry
records for several weeks or months and each message was transmitted several
times to ensure reception at the satellite. Bio-loggers were, typically,
deployed and programmed in a manner to maximize data transmission during the
spring pupping and molting period, though hourly percent-dry data were not
always successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (e.g., tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
atmospheric interference. Fortunately, missing records do not substantially bias
inference about haul-out probabilities [@conn2012c].

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). We calculated local solar
hour using the {solaR} package [@perpinan2012] within the R statistical
environment [@rcoreteam2021] based on the weighted daily average locations. We
also linked the weighted average daily locations to weather values from the
North American Regional Reanalysis (NARR) model produced by the National Centers
for Environmental Prediction [@mesinger2006a]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made available
across the region 8 times daily. For this study, NARR weather values were subset
to the extent of our study area over the Bering and Chukchi seas at 3-hr
intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003; @udevitz2009; @perry2017] and were
interpolated and assigned to daily seal locations using a bilinear method: 1)
air temperature at 2 m above the Earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the
Euclidean norm, 3) barometric pressure at sea level, and 4) precipitation 
(Table \@ref(tab:tblCovariates)).

For all seal species, we considered the following variables when modeling the
hourly haul-out behavior: day-of-year, solar hour, temperature, wind speed,
barometric pressure, precipitation, and wind chill (represented by a
*wind:temperature* interaction [@udevitz2009]). Ribbon and spotted seal models
included age-sex class and interactions between day-of-year and age-sex class,
but we omitted these from bearded seal models due to poor representation of
age-sex classes (Table \@ref(tab:tblDeploy)). Bearded seal models included a
latitudinal effect (and an interaction with day-of-year), since bearded seals
occupy a substantial range during the spring and we were interested in possible
differences in the timing of haul-out behavior along a latitudinal gradient. We
omitted the latitudinal effect from ribbon and spotted seal models because,
during the spring, these species are most prevalent near the southern ice edge
in the Bering Sea [@conn2013b].

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such, results are
shown for all ages (Figure \@ref(fig:beardedHOCal); see also 
\@ref(fig:beardedPredSE)). Additionally, after
approximately 9 months, 7 devices deployed on the rear flipper of bearded seals
reported implausible hourly percent-dry data (100% dry for several weeks but
indicative of movement and increasing transmission rates (see @boveng2013)). All
data after the first instance of unrealistic values were censored from this
analysis. Overall, bearded seals were less likely to haul out and had a bi-modal
distribution of haul-out probability across the day. In addition to a peak
around local solar noon, the bearded seal model predicted additional haul-out
activity around local midnight. In concert with the lower magnitude of haul-out
probability, bearded seal haul-out behavior was also more protracted throughout
the spring season compared to ribbon and spotted seals.

@@

When exploring the influence of weather, bearded seal haul-out probability
was strongly affected by wind
($F_{`r paste(bearded_type3$wind$Num.df, bearded_type3$wind$Den.df, sep=",")`}$
= `r bearded_type3$wind$F.value`; $p$ = `r bearded_type3$wind$Prob.F`)
and temperature
($F_{`r paste(bearded_type3$temp2m$Num.df, bearded_type3$temp2m$Den.df, sep=",")`}$
= `r bearded_type3$temp2m$F.value`; $p$ =
`r bearded_type3$temp2m$Prob.F`) with much higher haul-out probability during 
periods of higher temperatures and low wind speeds (Figure \@ref(fig:beardedHOwx)). 
Not surprisingly, wind chill 
($F_{`r paste(bearded_type3$temp2m_wind$Num.df, bearded_type3$temp2m_wind$Den.df, sep=",")`}$
= `r bearded_type3$temp2m_wind$F.value`; $p$ =
`r bearded_type3$temp2m_wind$Prob.F`) was also important. Barometric
pressure ($F_{`r paste(bearded_type3$pressure$Num.df, bearded_type3$pressure$Den.df, sep=",")`}$
= `r bearded_type3$pressure$F.value`; $p$ =
`r bearded_type3$pressure$Prob.F`) was also significant factor although less 
apparent (Figure \@ref(fig:beardedHOwx)). Any effect of precipitation 
was not a significant influence on haul-out probability 
($F_{`r paste(bearded_type3$precip$Num.df, bearded_type3$precip$Den.df, sep=",")`}$
= `r bearded_type3$precip$F.value`; $p$ =
`r bearded_type3$precip$Prob.F`).

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that was less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal);
see also \@ref(fig:spottedPredSE)). Adults of both sexes spent considerable time
in April, May and June hauled out. In contrast to ribbon seals, adult spotted
seal males had a more protracted haul-out season compared to females, and more
time out of the water in June (Figures \@ref(fig:ribbonHOCal) and
\@ref(fig:spottedHOCal)). As with ribbon seals, the young-of-the-year records
began after weaning and the model predictions demonstrated the development of
in-water activities (e.g. swimming, foraging) in May.

@@

Unlike previous analyses of seal haul-out behavior in spotted and ribbon seals
(e.g. [@verhoef2010a], [@conn2013b]), we also investigated the influence of
sex-age class on haul-out probabilities for both species (not for bearded seals
because of low sample size). Field identification of age class can be inexact,
particularly when differentiating subadults from adults. In the case of ribbon
seals, subadults often have less distinct ribbons than adults. Spotted seal
pelage cannot be used to reliably discern adults from subadults. Despite these
challenges, we feel the age classifications used in this analysis are useful in
testing for age-related effects on haul-out behavior.

@@

m2_ssrd_bearded <- mgcv::bam(
  dry ~ s(speno, bs = "re") + 
    era5_ssrd_watts +
    poly(day, 3, raw=TRUE) + 
    era5_ssrd_watts:poly(day, 3, raw=TRUE) +
    wind*temp2m + pressure + precip,
  data = bearded_model_data,
  family = binomial,
  AR.start = ar1_start,
  rho = lag1_bearded,
  discrete = TRUE)
```
